---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the \~ is equivalent of /Users/[youruser] on a Mac).


### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```


```{r}

#print will take a string and return it in the terminal
print("hello world")

```

------------------------------------------------

#pre_lab_01 notes#
#R basics#
# typing 2+2 into the console, then hitting enter will spit out number 4, function being it can compute equations and provide useful solutions. example seen below in console. however won't replicate into page, only console which is downside if want to continuously run and fix errors.#

ex.
```{r}
2+2
```

#variables can be assigned by using left facing arros <--, called assignment operator, can be seen in consol example. can have as many variables as can be named; reusable; anything can be stored --whole table, list of numbers, single word, whole book etc.#

```{r}
number <-- 2
```

```{r}
firstnumber <-- 1
secondnumber <-- 2
```

```{r}
(firstnumber + secondnumber) * secondnumber
```

#external libraries: "packages"; used to solve very specific problems -- example, Tidyverse. can install packages with function,, install.packages(),, and only have to do once

```{r}
install.packages('tidyverse')
```


#
In Rmarkdown data notebooks, code is written inside of codeblocks, and explanatory text in the white area outside of it.
```{r}
# This is a comment inside of a codeblock. Comments doesn't run when we run the codeblock.
# Adding this command is a good way to start your class reference notebook.
library(tidyverse)
```


#Aggregates#
#CSV: stripped down version of spreadsheet; each column is spearated by a comma#
#RDS files: less common when getting data from elsewhere/others#

RDS example: 
```{r}
umd_courses <- read_rds("data/umd_courses.rds")
```

#umd_courses: name of variable#
#read_rds(): function, only works when tidyverse is loaded; function takes parts of computer code that takes info and follows a series of pre-determined steps and shoots it back out#
#inside read_rds() is arguent: customize what function does; in this data set, each row represent course offered at UMD by departments during specific terms etc.#

#when use data sets/bases: looking into environment --> see number of rows ("obs" short for observations), number of columns (variables); can double click on dataframe name in envi. and explore like spreadsheet#

```{r}
glimpse(umd_courses)
```
#^^will provide list of columns, data type for each column and first few values for each column#

```{r}
head(umd_courses)
```
#^^print out columns and first six rows of data#

#package dplyr has function that can take massive amounts of data and analyze it; grouping like things together, doing simple things like counting them, averaging them together etc.#
#to do this, take dataset and introduce new operator |> "and then do this" type of function: called "pipe operator"; (shortcut for it: ctrl-shift-m)#
#can also use %>% --> same thing#

#data |> function: basically take the data set and then do this to it (commonly done/used)#

#very common pattern when using data sets:
data |> group_by(COLUMN NAME) |> summarize(VARIABLE NAME = AGGREGATE FUNCTION(COLUMN NAME))#

#in dataset, column with term information called "term"#

#code to count number of courses in each term:

```{r}
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n()
  )
```
dataset -- umd_course --
name -- term -- means group together all terms; which could be gotten by using glimpse()
after group data, need to count them up -- summarize()
inside summarize() needs to be the summaries one wants to find, hence putting -- count_classes = n()
-- 
count_classes: creating new field
setting equal to n(): fuction that counts number rows or records in each group#

#but this once run, aren't in order, so add |> arrange() to put data in order; default ascending order smallest - largest#

```{r}
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n()
  ) |>
  arrange(desc(count_classes))
```

#can even group more than one thing#

```{r}
umd_courses |>
  group_by(term, department) |>
  summarise(
    count_classes = n()
  ) |>
  arrange(term, department)
```

#sum(): know total number of data grouped; add up all the values in the column 
ex. to see total number of seats --> sum(seats)#
ex.
```{r}
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats)
  ) |>
  arrange(desc(total_seats))
```

#can also calculate average number of seats (the mean) and midpoint of data (the median)#
ex. 
```{r}
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats),
    mean_seats = mean(seats),
    median_seats = median(seats)
  ) |>
  arrange(desc(total_seats))
```

#highest and lowest points in data are, respectively, min() and max()#
ex.
```{r}
umd_courses |>
  group_by(department) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats),
    mean_seats = mean(seats),
    median_seats = median(seats),
    min_seats = min(seats),
    max_seats = max(seats)
  ) |>
  arrange(desc(total_seats))
```

#to sort from highest to lowest on a specific data point#
ex. 
```{r}
umd_courses |>
  arrange(desc(seats))
```


#n_designate can be used to help...


Chapter 18: Mutating Data
#can obtain percentages using -- dplyr and mutate -- to calcualte new metrics in new field using existing fields of data#

#essence of mutate --> using data you have to answer a new question#

#code to calculate percentages: with -summarize- n() is used to count things --> with mutate, a very similar syntax is used to calculate new value -- a new column of data -- using other values in dataset: to calculate percentage, need both the number and total number#

```{r}
general_22 <- general_22 |>
  mutate(
    total_votes = cox + moore + lashar + wallace + write_ins,
    pct_moore = moore/total_votes
  )
```

#however this makes a decimal expressed as a percentage --> to fix! multiply by 100: since replacing contents of new pct_moore column, can just update previous code and run it again: 

```{r}
general_22 <- general_22 |>
  mutate(
    pct_moore = (moore/total_votes)*100
  )
```

#now arragne#

```{r}
general_22 <- general_22 |>
  mutate(
    pct_moore = (moore/total_votes)*100
  ) |> 
  arrange(desc(pct_moore))
```

#now results ordered by pct_moore with highest % 1st; reverse is ofc with arrange function#

#mutate is also useful for standardizing data -- ex, making different spellings easier to comb through#

#when there is a mix of styles (lower-case and upper-case etc.) R will think of these as different names and try to organize the data accordingly: to fix! can use function -- str_to_upper -- which will convert character column into all uppercase: can be used for addresses, zip codes, anything that is misspelled

```{r}
standardized_maryland_expenses <- maryland_expenses |>
  mutate(
    payee_upper = str_to_upper(payee_name)
)
```


#mutate more powerful when combined with other functions# 

ex
```{r}
maryland_expenses_with_state <- maryland_expenses |>
  mutate(
    state = case_when(
        str_detect(address, " Maryland ") ~ "MD",
        str_detect(address, " California ") ~ "CA",
        str_detect(address, " Washington ") ~ "WA",
        str_detect(address, " Louisiana ") ~ "LA",
        str_detect(address, " Florida ") ~ "FL",
        str_detect(address, " North Carolina ") ~ "NC",
        str_detect(address, " Massachusetts ") ~ "MA",
        str_detect(address, " West Virginia ") ~ "WV",
        str_detect(address, " Virginia ") ~ "VA",
        .default = NA
      )
  )
```

#begins as typical mutate statement, but -- case_when -- introduces each line checks to see if pattern is contained in address column followed by a -- ~ -- and then a value for new column for records that match that check: can be read as: "if we find 'Maryland' in the address column, then put 'MD' in the state column": and if no match --> make state -- NA --#

#thus, new state column can be used to make summarizing easier#

```{r}
maryland_expenses_with_state |>
  group_by(state) |>
  summarize(total = sum(amount)) |>
  arrange(desc(total))
```


Chapter 17: Working with dates

#correct way to display dates: YYYY-MM-DD#

#problem so common that the tidyverse has an entire library for dealing with it -- lubridate -- to solve more common and less tricky problem

```{r}
install.packages('lubridate')
```

#example run through#

```{r}
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

```{r}
head(maryland_expenses)
```

#the first column --> looks like a date, but the -- <chr -- below the column name, means R thinks it's actually a character column: so need to make into a actual date column, which lubridate is good at doing; has variety of functions that match the format of the data#

#in this case, current format is m/d/y, and the lubridate function is called mdy that can be used with mutate#

```{r}
maryland_expenses <- maryland_expenses |> mutate(expenditure_date=mdy(expenditure_date))
```

```{r}
head(maryland_expenses)
```

#now R will recognize the column as a date column and looks like it should -- YYYY-MM-DD#

#lubridate has functions for almost any type of character date format, mdy, ymd, even datetimes like ymd_hms#

#the -- readr -- anticipates some date formating and can automatically handle many issues (indeed uses lubridate under its hood)#

#when importing CSV file, be sure to use read_csv, not read.csv#

#in spreadsheets can extract portions of dates --a month, day or year-- with formulas: can do same in R with lubridate

#Let’s say we wanted to add up the total amount spent in each month in our Maryland expenses data.

We could use formatting to create a Month field but that would group all the Aprils ever together. We could create a year and a month together, but that would give us an invalid date object and that would create problems later. Lubridate has something called a floor date that we can use.

So to follow along here, we’re going to use mutate to create a month field, group by to lump them together, summarize to count them up and arrange to order them. We’re just chaining things together.#

```{r}
maryland_expenses |>
  mutate(month = floor_date(expenditure_date, "month")) |>
  group_by(month) |>
  summarise(total_amount = sum(amount)) |>
  arrange(desc(total_amount))
```



Chapter 16: Filters and selections

#often have more data then necessary --> sometimes want to be rid of data#

#dplyr --> filtering and selecting#

#Filtering creates a subset of the data based on criteria. All records where the amount is greater than 150,000. All records that match “College Park”. Something like that. Filtering works with rows – when we filter, we get fewer rows back than we start with.#

#Selecting simply returns only the fields named. So if you only want to see city and amount, you select those fields. When you look at your data again, you’ll have two columns. If you try to use one of your columns that you had before you used select, you’ll get an error. Selecting works with columns. You will have the same number of records when you are done, but fewer columns of data to work with.#

#run through an example#

```{r}
library(tidyverse)
```

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
```


#if want to see only those courses offered by particular department, can use -- filter -- function to isolate just those records#

#filter works with something called a comparison operator: need to filter all records equal to "journalism": comparison operators in R, like most programing languages, are == for equal to, != for not equal to, > for greater than, >= for greater than or equal to and so on#

#!!!!!! -- be careful: = is not == -- and = is not "equal to"

= is an assignment operator in most languages - how things are named#

```{r}
journalism_courses <- umd_courses |> filter(department == "Journalism")

head(journalism_courses)
```

#if want to only work with course id and title...#

```{r}
selected_journalism_courses <- journalism_courses |> select(id, title)

head(selected_journalism_courses)
```


#if want to see only all the courses in theatre dept. with at least 15 seats#

#first way; chain together multiple filters#

```{r}
theatre_seats_15 <- umd_courses |> filter(department == "Theatre") |> filter(seats >= 15)

nrow(theatre_seats_15)
```

#gives up 308 records, but a single filter and boolean operators -- AND and OR -- can do this better#

#in this case, AND is & and OR is |#

#difference? with AND --> all conditions must be true to be included#

#difference? with OR --> any of those conditions things can be true and it will be included#

```{r}
and_theatre_seats_15 <- umd_courses |> filter(department == "Theatre" & seats >= 15)

nrow(and_theatre_seats_15)
```

#AND here gives the same answer as the attempt before#

```{r}
and_theatre_seats_15 <- umd_courses |> filter(department == "Theatre" | seats >= 15)

nrow(and_theatre_seats_15)
```

#OR gives 54,000 rows that are EITHER theatre classes OR have at least 15 seats#

#OR is additive; AND is restrictive#

#advice is to use filters one at a time and look at changes rather than doing multiple at once --> easier to see what is working this way#







