---
title: "Reverse Engineering Project"
author: "Fiona Roy"
date: "2025-03-12"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [CEOs gave heavily during legislative session, exposing loophole in NC's fundraising ban](https://www.wral.com/in-session-campaign-donations/17716104/)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}

# turn off sci notation
options(scipen=999)
library(tidyverse)
library(lubridate)
library(refinr)
library(janitor)
library(dplyr)

```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
# Load required data
# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")
donors <- read_csv("data/donors0103-0701.csv")

```


```{r}

donors <- donors |>
  mutate(
    amount = str_remove_all(amount, "\\$"),
    amount = str_remove_all(amount, ",")
  )

glimpse(donors)

```


```{r}
# Clean required data and prepare for analysis if needed. 

donors <- donors |>
  mutate(date=mdy(date)) |>
  mutate(amount = as.numeric(amount)) |>
  mutate(committee_zip = str_sub(committee_zip, start=1L, end=5L)) |>
  mutate(zip = str_sub(zip, start=1L, end=5L))

glimpse(donors)


```


## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: "Add in the week before session began, and the two businessmen gave more than $87,000 to Republican legislators and the North Carolina Republican Party, a key cog in the campaign machine this year as the two major parties fight for control of the General Assembly in November's elections."

* **Analysis summary**: The results of our analysis for this first sentence is that we were able to produce findings to support the claim that George Strobel and Robin Delmer gave more than 87,000 dollars throughout the North Carolina General Assembly. This then proves, at the very least, that the groundwork of data analysis for this article is off to a correct start (making the rest of the analysis, already, seem most likely to be true) in regard to its eventual, more specified, analysis of donations being given during session.

```{r}

donors <- donors |>
  rename(
    name_clean = "name-clean"
  )

```


```{r}

donors |>
  group_by(committee_name) |>
  summarize(
    amount = sum(amount)
  ) |>
  arrange(desc(amount))

```


```{r}
#date range of first week before NC general assembly --> 2018-01-03 to 2018-01-09##
#NC general assembly 2018 ran from 2018-01-10 to the 2018-07-04##
#looking for the two business men who gave more than 87,000 dollars to Republican party##
# Put code to reverse engineer sentence here
      
donors |>
  filter(name_clean == "GEORGE LEWIS STROBEL II" | name_clean == "ROBIN E DELMER") |>
  group_by(name_clean) |>
  summarise(amount = sum(amount)) |>
  arrange(desc(amount))
  
  
#name-clean, amount, date, committee_name, profession, employer_name, transaction_type#
# Display results of code below this codeblock

donors %>%
  filter(name_clean == "GEORGE LEWIS STROBEL II" | name_clean == "ROBIN E DELMER") %>%
  group_by(committee_name) %>%
  summarise(amount = sum(amount)) %>%
  arrange(desc(amount))
#rabon = dem
#michael lee = repub = 12500
#saine = repub = 12500
#hise = repub = 12500
#jackson = repub = 10000
#moore = repub = 10000
#brown = repub = 10000
#berger = repub = 10000
#nc rec = real estate commission = 6000
#randleman = repub = 5000
#szoka = repub = 4000
#brawley = repub = 2500
#lewis = repub = 2500 (later got in legal trouble for corruption lmao)
#tillman = repub = 2500
#bell = repub = 2500
#burr = repub = 1000


```

### Sentence 2

* **Sentence text**: "Senate President Pro Tem Phil Berger's campaign alone raised nearly $222,000 during session. Twenty-three people gave him the maximum allowed: $5,200."

* **Analysis summary**: We were able to confirm the finding when we filtered for Berger's campaign, the date's of the short session and the amount given (5,200) we got 23 rows. This number thus proves the sentence above that states 23 individual donators (either PACs or individual people) donated the max amount allowed, equaling $222,000.

```{r}
# Put code to reverse engineer sentence here
# Display results of code below this codeblock
#date range of first week before NC general assembly --> 2018-01-03 to 2018-01-09##
##legislative short session for NC general assembly may 16-june 29##
#NC general assembly 2018 ran from 2018-01-10 to the 2018-07-04##

berger <- donors |>
  filter(date >= '2018-05-16' & date <= '2018-06-29') |>
  filter(committee_name == "PHILIP E BERGER COMM") |>
  filter(amount == 5200) |>
  group_by(name_clean, date, amount) |>
  summarize(sum(amount))



```

### Sentence 3

* **Sentence text**: "Beer distributors also gave during session. Mark Craig, the president of distributor R.H. Barringer, put more than $30,000 into state legislative campaigns during the six-week period, including the maximum $5,200 to Berger. He was the top individual donor to legislative campaigns during session."

* **Analysis summary**: We were able to confirm that Mark Craig, put more than 33,000 dollars into the campaigns legislative session (from April 14, 2018, to June 28, 2018, he has donated 33200; although some donations were a little outside of the legislative window, I believe they count due to the whole loophole business). Additionally we were able to confirm that he did indeed also contribute a max amount donation of 5,200 dollars to Berger.

```{r}
# Put code to reverse engineer sentence here
# Display results of code below this codeblock

#have to run these block by block, but all answers are in here#

Mark_Craig <- donors |>
  filter(date >= '2018-05-16' & date <= '2018-06-29') |>
  filter(name_clean == "MARK R CRAIG") |>
  filter(amount == 5200) |>
  group_by(name_clean, date, amount) |>
  summarize(sum(amount))

##above proves the maximum 5200 donation to Berger##

donors |>
  filter(name_clean == "MARK R CRAIG") |>
  select(name_clean, date, amount) |>
  arrange(date)

##proves that his donations were all during the six-week legislative session##
##pulled the total_amount = sum(amount, na.rm = TRUE) part from chatgpt's advice to count the total amount of his donations##

donors |>
  filter(name_clean == "MARK R CRAIG") |>
  group_by(name_clean) |>
  summarise(
    total_amount = sum(amount, na.rm = TRUE)
  )
  
##proves that MARK R CRAIG has put more than 30000 dollars into the NC General Assembly session##

```

### Sentence 4

* **Sentence text**: "Add in donations to the two main political parties and to a handful of other large fundraising vehicles for state legislative races, and the in-session giving topped $3.4 million. Gov. Roy Cooper's main campaign account brought in another $407,000 during session."

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here
# Display results of code below this codeblock



```

### Sentence 5

* **Sentence text**: "A WRAL News analysis of campaign finance filings also turned up $16,000 in donations to legislative campaigns during session from 15 PACs with a lobbyist. Each would appear to be a violation of state fundraising rules, but spokespeople for most of the PACs, or the companies they represent, said there was a simple explanation: The checks were written before session."

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here
# Display results of code below this codeblock




```

-30-




