---
title: "pre_lab_08.Rmd"
author: "Derek Willis"
date: "2024-10-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 21

Visualizing data is becoming a much greater part of journalism. Large news organizations are creating graphics desks that create complex visuals with data to inform the public about important events.

To do it well is a course on its own. And not every story needs a feat of programming and art. Sometimes, you can help yourself and your story by just creating a quick chart, which helps you see patterns in the data that wouldn't otherwise surface.

Good news: one of the best libraries for visualizing data is in the tidyverse and it's pretty simple to make simple charts quickly with just a little bit of code. It's called [ggplot2](https://ggplot2.tidyverse.org/).

Let's revisit some data we've used in the past and turn it into charts. First, let's load libraries. When we load the tidyverse, we get ggplot2.

### Task 1: Load packages

**Task** Run the following code to load packages.

```{r}
library(tidyverse)
```

The dataset we'll use is voter registration data by county in Maryland from 2020 and September 2022. Let's load it.

### Task 2: Load data

**Task** Run the following code to load the data. In the space below, indicate how many rows and columns it has.

**Answer** 24 rows, 18 columns.

```{r}
md_voters <- read_csv("data/maryland_voters_2024_2022.csv")
```

## Bar charts

The first kind of chart we'll create is a simple bar chart.

It's a chart designed to show differences between things -- the magnitude of one thing, compared to the next thing, and the next, and the next.

So if we have thing, like a county, or a state, or a group name, and then a count of that group, we can make a bar chart.

So what does the chart of the top 10 maryland counties with the biggest change in unaffiliated voters from 2024 to 2022 look like?

First, we'll create a dataframe of those top 10, called maryland_top_counties.

### Task 3: Create a dataframe of top 10 counties for unaffiliated voter change in Maryland

**Task** Complete and run the following code to create a dataframe of top 10 counties with the biggest change in unaffiliated voters in Maryland. In the space below, explain what each step of the code below is doing in English. 

**Answer** The first part of the code is mutating the md_voters dataset to calculate a new column that calculates the difference in unaffiliated voters between 2022-2024. The next part of the code is making the calculations of the first part into a new dataset (for organizational and data purposes), arranging it so that the largest change in unaffiliated voters in MD counties is put first, and that the new dataset is organized by only the county and calculated difference (and only the top ten results). The last line of code just visually shows the new dataset in the r notebook rather than opening it up through the environment.

```{r}

md_voters <- md_voters |> 
  mutate(UNAF_DIFF = UNAF_2024 - UNAF_2022)

maryland_top_counties <- md_voters |>
  arrange(desc(UNAF_DIFF)) |>
  select(County, UNAF_DIFF) |> 
  head(10)

maryland_top_counties
```

Now let's create a bar chart using ggplot.

With ggplot, the first thing we'll always do is draw a blank canvas that will contain our chart. We start with our dataframe name, and then (%\>%) we invoke the ggplot() function to make that blank canvas. All this does is make a gray box, the blank canvas that will hold our chart.

### Task 4: Create an empty ggplot container

**Task** Run the following code. Explain what you see after running the code in the space below.

**Answer** A blank, grey box.

```{r}
maryland_top_counties |>
  ggplot()
```

Next we need to tell ggplot what kind of chart to make.

In ggplot, we work with two key concepts called geometries (abbreviated frequently as geom) and aesthetics (abbreviated as aes).

Geometries are the shape that the data will take; think of line charts, bar charts, scatterplots, histograms, pie charts and other common graphics forms.

Aesthetics help ggplot know what component of our data to visualize -- why we'll visualize values from one column instead of another.

In a bar chart, we first pass in the data to the geometry, then set the aesthetic.

In the codeblock below, we've added a new function, geom_bar().

Using geom_bar() -- as opposed to geom_line() -- says we're making a bar chart.

Inside of that function, the aesthetic, aes, says which columns to use in drawing the chart.

We're setting the values on the x axis (horizontal) to be the name of the county. We set weight to total loans, and it uses that value to "weight" or set the height of each bar.

One quirk here with ggplot.

After we've invoked the ggplot() function, you'll notice we're using a + symbol. It means the same thing as %\>% -- "and then do this". It's just a quirk of ggplot() that after you invoke the ggplot() function, you use + instead of %\>%. It makes no sense to me either, just something to live with.

### Task 5: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below.

**Answer** A bar chart with the new dataset about the unaffiliated differences, organized by each MD county (x-plot) and the actual numerical difference (y-plot) (used toe ggplot() function to do this).

```{r}
maryland_top_counties |>
  ggplot() +
  geom_bar(aes(x=County, weight=UNAF_DIFF))
```

This is a very basic chart. But it's hard to derive much meaning from this chart, because the counties aren't ordered from highest to lowest by total_loans. We can fix that by using the reorder() function to do just that:

### Task 6: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last codeblock?

**Answer** The bar chart is now organized by smallest to largest numerical difference so it is easier to read/visualize for readers. (it is reorder()[ed] to do this)

```{r}
maryland_top_counties |>
  ggplot() +
  geom_bar(aes(x=reorder(County,UNAF_DIFF), weight=UNAF_DIFF))
```

This is a little more useful. But the bottom is kind of a mess, with overlapping names. We can fix that by flipping it from a vertical bar chart (also called a column chart) to a horizontal one. coord_flip() does that for you.

### Task 7: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last codeblock?

**Answer** Now the bar chart has been flipped from vertical to a horizonal chart with the function coord_flip() to make it less messy and easier to read for viewers.

```{r}
maryland_top_counties |>
  ggplot() +
  geom_bar(aes(x=reorder(County,UNAF_DIFF), weight=UNAF_DIFF)) +
  coord_flip()
```

Is this art? No. Does it quickly tell you something meaningful? It does.

We're mainly going to use these charts to help us in reporting, so style isn't that important.

But it's worth mentioning that we can pretty up these charts for publication, if we wanted to, with some more code. To style the chart, we can change or even modify the "theme", a kind of skin that makes the chart look better.

It's kind of like applying CSS to html. Here I'm changing the theme slightly to remove the gray background with one of ggplot's built in themes, theme_minimal().

### Task 8: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last codeblock?

**Answer** Now the background of the chart is no longer grey, but white, to make it prettier (used the theme_minimal() code to do this.)

```{r}
maryland_top_counties |>
  ggplot() +
  geom_bar(aes(x=reorder(County,UNAF_DIFF), weight=UNAF_DIFF)) +
  coord_flip() + 
  theme_minimal()
```

The ggplot universe is pretty big, and lots of people have made and released cool themes for you to use. Want to make your graphics look kind of like [The Economist's](https://www.economist.com/) graphics? There's a theme for that.

First, you have to install and load a package that contains lots of extra themes, called [ggthemes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/).

### Task 9: Run the following code

**Task** Run the following code.

**Answer** don't know what kind of answer you want here, but ggthemes package is installed.

```{r}
install.packages('ggthemes')
library(ggthemes)
```

And now we'll apply the economist theme from that package with theme_economist()

### Task 10: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last version of the chart?

**Answer** Now the whole chart's background is a blue/grey color, and the bars are laid on top of white lines to make it more symmetrical and organized. The fonts is also darker and more simplistic to read (black, bolded [to a degree], and slightly larger in size; more aligned with bars on chart). (done with theme_economist()).

```{r}
maryland_top_counties |>
  ggplot() +
  geom_bar(aes(x=reorder(County,UNAF_DIFF), weight=UNAF_DIFF)) +
  coord_flip() + 
  theme_economist()
```

Those axis titles are kind of a mess. Let's change "count" on the x axis to "net change" and change "reorder(County,TOTAL_DIFF)" to "county". And while we're at it, let's add a basic title and a source as a caption. We'll use a new function, labs(), which is short for labels.

### Task 11: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last codeblock?

**Answer** Axis titles have been changed to x axis --> net change; y axis --> county. There is also now a title along the top, "Counties with Biggest Jumps in Unaffiliated Voters", and source title along the bottom of the graph, "source: Maryland State Board of Elections". Although the spacing needs to be better. (done with labs() function).

```{r}
maryland_top_counties |>
  ggplot() +
  geom_bar(aes(x=reorder(County,UNAF_DIFF), weight=UNAF_DIFF)) +
  coord_flip() + 
  theme_economist() +
  labs(
    title="Counties with Biggest Jumps in Unaffiliated Voters",
    x = "net change",
    y = "county",
    caption = "source: Maryland State Board of Elections"
    
  )
```

Viola. Not super pretty, but good enough to show an editor to help them understand the conclusions you reached with your data analysis.

## Line charts

Let's look at how to make another common chart type that will help you understand patterns in your data.

Line charts can show change over time. It works much the same as a bar chart, code wise, but instead of a weight, it uses a y.

So, let's load some WinRed contribution data we've previously used and create a dataframe with a count of contributions for each date in our data.

### Task 12: Load the WinRed contribution data and create a dataframe of contributions by date

**Task** Run the following code to load the WinRed data

```{r}
md_winred <- read_rds("data/maryland_winred.rds")
```

**Task** Run the following code. Explain what you see after running the code in the space below.

**Answer** After running this code, I see below a new dataset that organizes md_winred info into total contributions/donations (by all people) entered per date entered into the original dataset. (how many people/organizations/etc. donated per day included in the original dataset)

```{r}

md_winred_by_date <- md_winred |>
  group_by(date) |>
  summarise(
    total_contributions=n()
  )

md_winred_by_date 
```

And now let's make a line chart to look for patterns in this data.

We'll put the date on the x axis and total contributions on the y axis.

### Task 13: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below.

**Answer** A basic line chart that has the date on the x-axis and the total number of contributions on the y-axis.

```{r}
md_winred_by_date |>
  ggplot() + 
  geom_line(aes(x=date, y=total_contributions))


```

It's not super pretty, but there's a pattern here: the number of contributions fluctuates between 400 and 800 a day for most of this period, and then jumps way up at the end of June. We've learned that the end of June is the end of a reporting period, and donors respond to deadlines.

Right now, it's kind of hard to see specifics, though. When did some of those smaller spikes and troughs happen?

We can't really tell. So let's modify the x axis to have one tick mark and label per month. We can do that with a function called scale_x_date().

We'll set the date_breaks to appear for every week; if we wanted every month, we'd say date_breaks = "1 month". We can set the date to appear as month abbreviated name (%b) and day (%d).

### Task 14: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below.

**Answer** Now the line chart has more specifications in it: the x axis now has one tick mark and label per month to indicate the dates better (done with scale_x_date()). These date_breaks appear every week.

```{r}
md_winred_by_date |>
  ggplot() + 
  geom_line(aes(x=date, y=total_contributions)) + 
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d")


```

Those are a little hard to read, so we can turn them 45 degrees to remove the overlap using the theme() function for styling. With "axis.text.x = element_text(angle = 45, hjust=1)" we're saying, turn the date labels 45 degrees.

### Task 15: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below.

**Answer** To make it easier for reader comprehension, the new date titles along the x-axis have been turned 45 degrees to remove overlap using the theme() function for styling. 

```{r}
md_winred_by_date |>
  ggplot() + 
  geom_line(aes(x=date, y=total_contributions)) + 
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d") +
  theme(
    axis.text.x = element_text(angle = 45,  hjust=1)
  )

```

Again, this isn't as pretty as we could make it. But by charting this, we can quickly see a pattern that can help guide our reporting.

We're just scratching the surface of what ggplot can do, and chart types. There's so much more you can do, so many other chart types you can make. But the basics we've shown here will get you started.




