---
title: "lab_01"
author: "Ryan Little"
date: "2025-02-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your reference notebook!

## Load libraries and establish settings

**Task** Import the Tidyverse. Write the code needed to load the library. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
library(tidyverse)
```

## Load Data

In this lab, we will be recreating the first data story I ever did as a data journalism student at Merrill while working in the Howard Center for Investigative Journalism. First, we need to import the court charges data from Okaloosa County, Florida.

**Task** Using the example from the pre_lab, write code to load the CSV file in the data file and assign it to a good variable name. This data is only charges for homeless people.

```{r}
okaloosa_court_charges <- read_csv("data/okaloosa-court-charges.csv")
```

## Examine the data

**Task** Run the codeblock below to use glimpse to get a sense of the column names, types and sample data.

**Q1** How many rows and columns does this dataframe have?

**A1** There are 15,386 rows and 56 columns.

```{r}
glimpse(okaloosa_court_charges)
```

## Answer questions

**Q2** Which charges are homeless people most often charged with?

**A2** Homeless people are most often charged with narcotic equip possess or use.

```{r}
 okaloosa_court_charges |>
  group_by(court_statute) |>
  summarise(
    count_contribs = n()
  ) |>
    arrange(desc(count_contribs))
  
```

**Q3** What charges against homeless people have been charged the least?

**A3** The charges that homeless people have least been charged with are adult permits under 16 yoa to possess weapon.

```{r}
 okaloosa_court_charges |>
  group_by(court_statute) |>
  summarise(
    count_contribs = n()
  ) |>
    arrange((count_contribs))
```

**Q4** Why do you think there are so many charges identified in **A3** that are tied for the least common? Hint: You can't answer this question with code.

**A4** I think there are so many charges identified in the question before that are tied for least common because these are all charges that are not common and/or are extremely unlikely things for most people to be commonly doing/being charged with. In other words, there are so many individual cases of very specific crimes, becuase these crimes are not ones most common people would consider committing. Additionally, it could be due to human error in recording certain crimes -- i.e. spelling mistakes.

**Q5** Which prosecutor has prosecuted the most charges against homeless people? Is the top prosecutor related to any famous Floridians?

**A5** The prosecutor that has prosecuted the most charges against homeless people is Ginger Bowden Madden. She is related to the famous Floridians Bobby Bowden.

```{r}
 okaloosa_court_charges |>
  group_by(prosecutor) |>
  summarise(
    count_contribs = n()
  ) |>
    arrange(desc(count_contribs))
```


**Q6** Which judge has overseen the most cases involving homeless people? Hint: Cases and charges are not the same thing.

**A6** The judge who has overseen the most cases invovleing homeless people is Patricia s Grinsted.

```{r}
 okaloosa_court_charges |>
  group_by(judge) |>
  summarise(
    count_contribs = n_distinct(case_number)
  ) |>
    arrange(desc(count_contribs))
```

**Q7** What prosecutor has prosecuted the most cases against a single homeless person?

**A7** The prosecutor who has prosecuted the most cases against a single homeless person is Cassie Reed.

```{r}
okaloosa_court_charges |>
  group_by(prosecutor, defendant) |>
  summarise(
    count_contribs = n()
  ) |>
    arrange(desc(count_contribs))
```

**Q8** We need to create a dataframe that will help us when we are reporting on the ground. Building off our work so far, we need a dataframe that includes the name of every homeless person who was charged with a crime and some important counts about their interactions with the criminal justice system. Create a dataframe that includes the name of each homeless person, the number of charges they have faced, the number of cases they have been involved in, how many dollars they have been assessed in fines, the average fine imposed on them and the median fine imposed upon them. Then sort the data frame so the person with the most charges is at the top. 

**A8** The person with the most charges is Richard Scott Chancey.

```{r}
okaloosa_court_charges |>
  group_by(defendant) |>
  summarise(
    charges = n(), 
    number_cases = n_distinct(case_number),
    total_fine = sum(total_assessed),
    avg_fine = mean(total_assessed),
    median_fine = median(total_assessed)
  ) |>
    arrange(desc(charges))
```

**Q9** Please describe a short reporting plan for how you would use the data you have generated in **A8** to report on the criminal justice system in Okaloosa County, Florida. Who are the people we should talk to? What more should we do with our data analysis to be certain we know what the story is?

**A9** A short reporting plan that utilizes this data, which is focused on the criminal justice system in Okaloosa County, Florida, could be to actively search for, and talk to, the ten people on the list who have the greatest number of charges and fines against them. By focusing on the people who have the most charges and fines against them we can begin to understand what the most common charges against homeless people are in Okaloosa County, and how they are expected to serve for their 'crime' (i.e. figure out most common charges, and figure out why fines are commonly assigned and how they expect the homeless to pay such expensive fines). Additionally, I would try to gather a comment from the police force that is issuing these charges, and the courts which are assigning the fines, in order to understand their reasoning. 
What we should do with our data analysis to be certain that we know what the story is is to run it more than one time (to weed out errors), as well as confer with both the police/courts (the place where the data was provided by) as well as the homeless people, the charges and fines are against, to confirm that the data we have (and more importantly are using) is correct.

