---
title: "lab_04"
author: "Derek Willis"
date: "2024-09-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(janitor)
library(lubridate)

```

For this lab, we want to investigate spending by the two leading Senate candidates in Maryland during this election, Angela Alsobrooks and Larry Hogan. Our goal is to try and standardize the purpose of the expenditures so that we can compare the two campaigns. We have the data, but it's messy - purpose descriptions are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of campaign expenditures located in the data folder called "md_senate_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table

# Load data
md_senate_expenses <- read_csv("data/md_senate_expenses.csv")

# Display it; date column already listed as date
glimpse(md_senate_expenses)

```

## Answer questions

**Q1.** You've been assigned to compare the spending priorities of the two campaigns, but they describe their spending differently using the `purpose` column. Using OpenRefine, create a project using the original CSV file, make a copy of the `purpose` column called `purpose_clean` and then standardize it, focusing on the purposes that appear most often or have the largest total amounts. You can transform the data as you want - making all records the same case, for example, to help. The goal is to be able to group together as many similar purpose values, so you can choose to call all transportation-related expenses the same thing.

How do you know when you're done? It's a judgment call, but you should be close to 100 unique values for `purpose_clean`, and maybe even below that number.

Then, when you are done standardizing `purpose_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.

Then, using that `purpose_clean` column, write code to find the largest amounts of spending on categories by both campaigns.

**A1.** It appears that the largest amounts of spending on categories by both campaigns are in regard to advertising: Alsobrooks for media advertising (3639344.50), and Hogan for ADS (2178145.48).

```{r}
# Load data
clean_md_senate_expenses <- read_csv("data/cleaned-md-senate-expenses-csv.csv")

# largest amounts of spending on categories in both campaigns --> purpose_clean

clean_md_senate_expenses |>
  group_by(purpose_clean, candidate) |>
  summarize(
    largest_amount= sum(amount)
  ) |>
  arrange(desc(largest_amount))


```

**Q2.** Let's do some research on the top purposes from A1. Feel free to search the Web to understand those terms better if need be, and then write a description comparing and contrasting the two campaigns' largest categories of spending.

**A2.** Both candidates campaigns seem to focus the majority of their spending on advertising and staffing. Alsobrooks campaign seems to focus most on media advertising, salary for campaign members, and list aquisition (gathering lists of info of potential donors, fundraisers, vendors etc.). Hogan's campaign focuses most of his revenue on ADS (although he focuses more on physical, printed adverts compared to Alsobrooks who favors online/media ads) and staffing for his campaign. 


**Q3.** Pick one of the top categories for each campaign and write code that shows how that spending has changed over time, if at all. You'll need to use lubridate for this. Write a couple of sentences describing that pattern.

**A3.** For Alsobrooks, when focusing on her top spending category -- Media Advertising -- it seems like each week she (pretty consistantly except for the week of 2024-03-17) steadily increases the amount of money she spends. For Hogan, when focusing on his top spending category -- ADS -- it appears that there is no steady incline or decline of his spending. Each week for Hogan is different, spending amounts differing each week drastically, like jumping from just under 3,000 one week to a little over 300,000 the next. 

```{r}

# Alsobrooks focus on media ads
clean_md_senate_expenses |>
  mutate(week = floor_date(date, "week")) |>
  filter(purpose_clean =="Media Advertising", candidate == "Alsobrooks") |>
  group_by(week, candidate) |>
  summarize(
    spending_amount=sum(amount)
  ) |>
  arrange(week)

# Hogan focus on ADS
clean_md_senate_expenses |>
  mutate(week = floor_date(date, "week")) |>
  filter(purpose_clean =="ADS", candidate == "Hogan") |>
  group_by(week, candidate) |>
  summarize(
    spending_amount=sum(amount)
  ) |>
  arrange(week)


```


**Q4.** Which committee has the larger payroll or salary spending so far? How many unique recipients of payroll/staff spending are there for each candidate?

**A4.** The committee that has the larger payroll/salary spending so far is Hogan's campaign, although Alsobrooks has the more individual unique recipients (Hogan = total spending of 481109.5; one unique recipient)(Alsobrooks = total spending of 268885.30 for salary [369376.50 if including payroll tax and fees]; 22 unique salary recipients [24 if including payroll tax and fees]).

```{r}

# Alsobrooks; used ChatGPT for filter advice on this one
clean_md_senate_expenses |>
  filter(purpose_clean == c("Payroll Fee","Payroll Taxes", "Salary"), candidate == "Alsobrooks") |>
  group_by(purpose_clean, candidate) |>
  summarize(
    total_spending = sum(amount),
    unique_recipients = n_distinct(first_name, last_name)
  ) |>
  arrange(desc(total_spending))

#Hogan; unsure about this one due to his confusing staffing records, this is all I could think of to include
clean_md_senate_expenses |>
  filter(purpose_clean == "STAFFING", candidate == "Hogan") |>
  group_by(purpose_clean, candidate) |>
  summarize(
    total_spending = sum(amount),
    unique_recipients = n_distinct(first_name, last_name, organization_name)
  ) |>
  arrange(desc(total_spending))

```

```{r}

#this is the Q4 answer with the additional filters jic
# Alsobrooks; used ChatGPT for filter advice on this one
clean_md_senate_expenses |>
  filter(purpose_clean == c("Payroll Fee","Payroll Taxes", "Salary", "Stipend", "Travel Stipend", "Research Consulting", "Legal Consulting", "Digital Consulting", "Compliance Consulting", "Fundraising Consulting", "Communication Consulting", "STRATEGY CONSULTING", "Campaign Management Consulting", "General Strategy Consulting"), candidate == "Alsobrooks") |>
  group_by(purpose_clean, candidate) |>
  summarize(
    total_spending = sum(amount),
    unique_recipients = n_distinct(first_name, last_name)
  ) |>
  arrange(desc(total_spending))

#Hogan; unsure about this one due to his confusing staffing records, this is all I could think of to include
clean_md_senate_expenses |>
  filter(purpose_clean == c("STAFFING", "STRATEGY CONSULTING", "Travel Stipend", "CONTRACTOR FIELD STAFF"), candidate == "Hogan") |>
  group_by(purpose_clean, candidate) |>
  summarize(
    total_spending = sum(amount),
    unique_recipients = n_distinct(first_name, last_name, organization_name)
  ) |>
  arrange(desc(total_spending))

```


**Q5** What's the most interesting story idea you've found in this data?

**A5** One interesting story idea that I've found in this data is the drastic jumping/differences in Hogans top spending category, ADS, and how the numbers are so widely different each week. I want to know why this is, what specifically were his motivations for the difference in spending each week etc. as in comparison, Alsobrooks followed a pretty steady increase in spending on her preferred method of advert use. 



