---
title: "lab_09.Rmd"
author: "derek willis"
date: "2024-10-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Datawrapper account
-   Election data in CSV files in the `data` folder
-   Tidycensus

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Turn off scientific notation
options(scipen=999)

# load libraries here
library(tidycensus)
library(janitor)
```

## Questions

**Q1.** You're interested in seeing how the methods of voting have changed in Maryland since 2018. Using the `md_votes_by_type.csv` that has data from the top of the ticket (governor in 2018 and 2022, president in 2020), make a chart showing the percentage of votes cast in each category - early voting, election day and mail - over the three years. You'll need to make those percentages in R and then export the CSV using `write_csv`, then make the chart in Datawrapper. Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect.

HINT: In Datawrapper's "Check and Describe" tab, you can click on a column to get the option of "hiding" it in any visualization.

What do you think the story here is? Put the answer below along with the URL of your chart.

**A1.** (LINK: https://datawrapper.dwcdn.net/16HLK/2/ ) The story here should focus on why for Maryland Governor races do constituents prefer to vote on election day, but for presidential races, Maryland constituents prefer to vote by mail. Is it a matter of governor races being more local? Or is there another reason? Could it be because of the times in which the election day, in-person, voting slots are held are easier to get to by the average constituent for governor races rather than presidential races?

```{r}
##uploading csv file##
md_vote_types <- read_csv("data/md_votes_by_type.csv")

md_vote_types
```

```{r}
##fixing up the csv file, making it pct out of total##

md_vote_types <- md_vote_types |>
  mutate(Early = (Early/Total)*100) |>
  mutate(Election_Day = (Election_Day/Total)*100) |>
  mutate(Mail = (Mail/Total)*100)

write_csv(md_vote_types, "data/md_vote_types.csv")
```


**Q2.** You've been assigned to track early-voting turnout in Maryland at the county level. To do that, use [data from the state](https://elections.maryland.gov/press_room/2024_stats/PG24/EarlyVoting%20RAW%20data.csv). YOU DO NOT NEED TO LOAD THIS FILE. I've created several CSV files from it; each has a row for each day, with the counties as columns, so you can track early voting turnout in certain categories. The available files in the data folder are:

* ev_totals_18_24.csv - early voting turnout by the 18-24 age group
* ev_totals_65_up.csv - early voting turnout by the 65+ age group
* ev_totals_rep.csv - early voting turnout by Republicans
* ev_totals_dem.csv - early voting turnout by Democrats
* ev_totals_una.csv - early voting turnout by unaffiliated voters
* ev_totals_women.csv - early voting turnout by women
* ev_totals_men.csv - early voting turnout by men

Which file you use to make your chart, and what kind of chart is up to you (but it MUST be a chart, not a map or table), so experiment to see what works best. Since we have data over time, you might start with a line chart, but you might not end up there. You can display all counties, or just some of them - to limit the number of counties, you'll need to write code to select only the columns that you need and then use `write_csv()` to export the dataframe to a CSV file that you'll use in Datawrapper.

Your chart MUST: 

\* Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect. 
\* Easily readable labels and values 
\* Be published
\* Put the question you chose to answer below along, with the URL of your chart.

Tell me below why you picked the CSV you did. If you made any changes to the data, put that code below.

**A2.** (LINK: https://datawrapper.dwcdn.net/F58Kl/1/) I chose this CSV because I thought it would be interesting to see the difference between counties in regard to which ones had the most versus the least male voters - a demographic that has seen a sort of decline (in some areas) over the past few years. The question I chose to answer/show in my chart was: Which Maryland counties had the highest male early voter turn out rate where each day received over a minimum of 1,000 votes?

```{r}
##early voting turn out by men##
##uploading csv for a quick little looksie##

ev_totals_men <- read_csv("data/ev_totals_men.csv")

##looked through data --> too much to put all in one chart unfortunately :( --> going to make chart based off of the counties that had a turn out of 1,000+ votes for all six days to narrow down; so counties with highest early voter turn outs - focused on those who identified as men##
ev_totals_men <- ev_totals_men |>
  select(Day, `Anne Arundel`, `Baltimore City`, `Baltimore County`, Carroll, Charles, Frederick, Harford, Howard, Montgomery, `Prince George's`)

ev_totals_men

write_csv(ev_totals_men, "data/ev_totals_men.csv")
```


**Q3.** After reading the recent CNN story about fundraising called "[How elderly dementia patients are unwittingly fueling political campaigns](https://www.cnn.com/interactive/2024/10/politics/political-fundraising-elderly-election-invs-dg/)," you decide to dig into the habits of Maryland political donors who give money via ActBlue and WinRed.

You want to show Maryland's political geography - where those donors live - and the frequency of donations. To do that, you'll load in two CSV files: `md_winred_zip_totals.csv` and `md_actblue_zip_totals.csv`.

You also will need to get population by zip code for Maryland - we did a similar thing in a previous lab, but this time you want the total population as your variable. Use tidycensus to do that, and save that dataframe, then join it to each of your contribution dataframes - you will need to make sure that the zip codes you're joining on are the same datatype.

In each dataframe, use the _number of donations_ to calculate a per capita rate of the number of donations per 1,000 population for each zip code using `mutate()`. Calculating a rate works like this: it's the thing you're counting divided by the population times the "per" number.

Then export each dataframe using `write_csv` and make two (2) choropleth maps in Datawrapper. For each map, you will need to:

-   Search for Maryland zip codes (use the 2020 data) as the base map
-   Upload the CSV file
-   Change the tooltip to show both the rate of donations and the total amount (you can put them side by side or put one under the other, as we did in the pre_lab). Make sure to properly format both numbers to make them readable.
-   Have a title, description and source that tells readers about the data (the source is the Federal Election Commission, and the data covers Jan. 1, 2023 through Oct. 16, 2024). The title and description should highlight the most newsworthy or important aspect.

Compare the two maps. What do you think are the most interesting story ideas in both, and what would you want to know more about? Put the answer below along with the URLs of your maps.

**A3.** (LINK WinRed: https://datawrapper.dwcdn.net/gqlYX/2/) (LINK ActBlue: https://datawrapper.dwcdn.net/ruaJ4/1/) I think that the most interesting story idea for the WinRed map is that, according to the data that I used, there is only one zipcode that is even remotely blue (the color of the gradient I chose that shows the highest levels of donations), which is Hagerstown, MD. This level of donation compared to the rest of Maryland for WinRed fundraising was very shocking to me as it stood out so much. I think I would focus my story on this map about why there is such a drastic difference because I want to know why this is: Is the data wrong? Is there something in particular in this ZIP that is attracting these donations? or something else along those lines. In regard to the ActBlue map, what is most interesting (although not at all surprising considering my knowledge of Maryland's most Democrat sectors) is that the highest donations by ZIP are centered around where D.C. is and where Baltimore City is. Thus, I would center my story around the opposite, I would want to focus on the ZIP's that (as shown by the data I obtained) have a decent amount of donations, but are in mostly Republican central areas of Maryland, and find out when this change began occurring. 

```{r}
##uploading csv file##
md_winred_zip_totals <- read_csv("data/md_winred_zip_totals.csv")

##uploading csv file##
md_actblue_zip_totals <- read_csv("data/md_actblue_zip_totals.csv")
```

```{r}
##getting census population by zip code for MD data; total pop. as variable##

census_population_zip <- get_acs(geography = "zcta", variable = "B01001_001", year = 2020, survey = "acs5") |>
  rename(zip = GEOID) |>
  mutate(zip = as.integer(zip))

```

```{r}
##joining the census with the csv files##

md_winred_zip_totals <- md_winred_zip_totals |> left_join(census_population_zip)

md_actblue_zip_totals <- md_actblue_zip_totals |> left_join(census_population_zip)
```

```{r}
##calculating per capita##

md_winred_zip_totals <- md_winred_zip_totals |>
  mutate(per_capita_1k = (donations/estimate)*1000)

md_actblue_zip_totals <- md_actblue_zip_totals |>
  mutate(per_capita_1k = (donations/estimate)*1000)

```

```{r}
##write_csv to establish changes to og file##
write_csv(md_winred_zip_totals, "data/md_winred_zip_totals.csv")

write_csv(md_actblue_zip_totals, "data/md_actblue_zip_totals.csv")
```


